---
title: "Easy Exercise: Import Your Own Corpus"
subtitle: "Text Analysis with R"
format:
  html:
    toc: true
    toc-depth: 2
execute:
  echo: true
  warning: false
  message: false
---


## Aim

In this exercise you will:

- put your own `.txt` files in one folder
- import them into R
- split them into sentences
- split them into word tokens
- remove stopwords

This is a simple starting workflow. It is designed to work with plain text files.

## Before You Start

Create a folder called `my_corpus` inside this project folder.

Put your own `.txt` files inside it.

Example:

```text
my_corpus/
  text_1.txt
  text_2.txt
  text_3.txt
```

Try to use short, simple filenames with no spaces.

Good filenames:

```text
austen_pride.txt
poe_reddeath.txt
my_interview_01.txt
```

Avoid filenames like:

```text
my text final version!!!.txt
```

## Setup

```{r}
# List the packages we need for this exercise.
required_packages <- c("tidyverse", "readtext", "tidytext")

# Check which packages are not installed yet.
missing_packages <- required_packages[!vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)]

# Install only the missing packages.
if (length(missing_packages) > 0) {
  install.packages(missing_packages)
}

# Load the packages so we can use their functions.
library(tidyverse)
library(readtext)
library(tidytext)
```

## Check Your Folder

```{r}
# This is the folder where your text files should be.
corpus_folder <- "my_corpus"

# Check whether the folder exists.
dir.exists(corpus_folder)
```

If the result is `FALSE`, create the folder `my_corpus` in the project folder and put your `.txt` files inside it.

```{r}
# Find all .txt files inside the folder.
text_files <- list.files(
  corpus_folder,
  pattern = "\\.txt$",
  full.names = TRUE
)

# Show the files R found.
text_files
```

```{r}
# Stop with a clear message if R did not find any .txt files.
if (length(text_files) == 0) {
  stop("No .txt files found. Put your plain text files inside the my_corpus folder.")
}
```

## Import Your Text Files

```{r}
# Import all .txt files found in the folder.
# Each file becomes one row in the data frame.
corpus_docs <- readtext(text_files, encoding = "UTF-8") %>%
  # Convert the result to a tibble, which prints nicely in R.
  as_tibble()

# Show the imported documents.
corpus_docs
```

## Check the Import

```{r}
# Count how many documents were imported.
nrow(corpus_docs)
```

```{r}
# Show the document IDs.
# These usually come from the filenames.
corpus_docs %>%
  select(doc_id)
```

```{r}
# Count the number of characters in each document.
# This is a quick way to spot empty or unexpectedly short files.
corpus_docs %>%
  mutate(n_characters = str_length(text)) %>%
  select(doc_id, n_characters)
```

## Clean the Document Text

```{r}
# Clean the document-level text.
corpus_docs <- corpus_docs %>%
  # Replace repeated spaces, tabs, and line breaks with one space.
  mutate(text = str_squish(text))

# Check the cleaned documents.
corpus_docs
```

## Split Into Sentences

```{r}
# Create a sentence-level corpus.
corpus_sentences <- corpus_docs %>%
  # Split each document into sentences.
  unnest_sentences(
    output = "sentence",
    input = "text",
    to_lower = FALSE
  ) %>%
  # Remove extra spaces inside each sentence.
  mutate(sentence = str_squish(sentence)) %>%
  # Group by document so sentence numbers restart for each file.
  group_by(doc_id) %>%
  # Add a sentence ID within each document.
  mutate(sentence_id = row_number()) %>%
  # Ungroup so later steps work on the full table.
  ungroup()

# Show the first sentences.
head(corpus_sentences)
```

```{r}
# Count how many sentences each document has.
corpus_sentences %>%
  count(doc_id, name = "n_sentences")
```

## Split Into Word Tokens

```{r}
# Create a token-level corpus.
corpus_tokens <- corpus_sentences %>%
  # Split each sentence into words.
  unnest_tokens(
    output = "word",
    input = "sentence",
    to_lower = TRUE,
    drop = FALSE
  ) %>%
  # Group by document and sentence so token numbers restart in each sentence.
  group_by(doc_id, sentence_id) %>%
  # Add a token ID to keep word order.
  mutate(token_id = row_number()) %>%
  # Ungroup so later steps work on the full table.
  ungroup()

# Show the first tokens.
head(corpus_tokens, 20)
```

```{r}
# Count how many word tokens each document has.
corpus_tokens %>%
  count(doc_id, name = "n_words")
```

## Remove Stopwords

Stopwords are very common words such as `the`, `and`, `of`, and `to`.

Removing stopwords can make word frequency lists easier to read, but it also removes part of the language. Always think about whether stopword removal makes sense for your question.

```{r}
# Look at the stopword list from tidytext.
head(stop_words)
```

```{r}
# Remove stopwords from the token table.
corpus_tokens_no_stopwords <- corpus_tokens %>%
  # anti_join keeps rows from corpus_tokens where the word is NOT in stop_words.
  anti_join(stop_words, by = "word")

# Show the first tokens after stopword removal.
head(corpus_tokens_no_stopwords, 20)
```

```{r}
# Compare the number of tokens before and after stopword removal.
tibble(
  step = c("before stopword removal", "after stopword removal"),
  n_tokens = c(nrow(corpus_tokens), nrow(corpus_tokens_no_stopwords))
)
```

## Find Common Words

```{r}
# Count the most common words after stopword removal.
common_words <- corpus_tokens_no_stopwords %>%
  count(word, sort = TRUE)

# Show the top 20 words.
common_words %>%
  slice_head(n = 20)
```

```{r}
# Count the most common words in each document.
common_words_by_document <- corpus_tokens_no_stopwords %>%
  count(doc_id, word, sort = TRUE)

# Show the first rows.
common_words_by_document %>%
  group_by(doc_id) %>%
  slice_max(n, n = 10) %>%
  ungroup()
```

## Save Your Corpus Objects

This step is optional, but useful. It saves your imported corpus so you can use it again later.

```{r}
# Save the document-level corpus.
save(corpus_docs, file = "my_corpus_docs.RData")

# Save the sentence-level corpus.
save(corpus_sentences, file = "my_corpus_sentences.RData")

# Save the token-level corpus without stopwords.
save(corpus_tokens_no_stopwords, file = "my_corpus_tokens_no_stopwords.RData")
```

## Reflection

Before analysing the results, ask:

- Which texts did I include?
- Which texts did I leave out?
- Are my files complete and clean?
- Does sentence splitting look correct?
- Does stopword removal help my question, or hide something important?
- What should I check by reading the original sentences?
