---
title: "01. Introduction to R for Text Analysis"
subtitle: "Text Analysis with R"
format:
  html:
    toc: true
    toc-depth: 2
    number-sections: true
execute:
  echo: true
  warning: false
  message: false
---

## Aim

This notebook introduces the basic R objects and operations we will use in the course. We will create variables, access data, read text files, write simple output, and load packages.

In RStudio or Posit Cloud, open `UoM_Digital_Methods_TM.Rproj` before running the notebook. This keeps file paths consistent.

## Creating Variables

Variables are containers for values. In R, we usually create them with `<-`. Variables can hold different types of data, such as logical values, numbers, and strings.

```{r}
# Create a logical variable.
# Logical values can only be TRUE or FALSE.
my_logical <- TRUE

# Print the value of the object we just created.
my_logical
```

```{r}
# Create a numeric variable.
# Numeric values can be used for arithmetic.
my_number <- 1

# Print the number.
my_number
```

```{r}
# Create a character variable.
# Character values contain text and must be placed inside quotation marks.
my_string <- "to be or not to be"

# Print the string.
my_string
```

Vectors hold several values of the same general type.

```{r}
# Create a numeric vector using c(), which means "combine".
my_first_vector <- c(1, 2, 3, 4, 5)
my_first_vector

# Replace the same object with a shorter way to create the sequence 1 to 5.
my_first_vector <- 1:5
my_first_vector
```

```{r}
# Create a second vector, this time with character values.
# Each word is one element of the vector.
my_second_vector <- c("to", "be", "or", "not", "to", "be")
my_second_vector
```

Lists can hold different kinds of objects.

```{r}
# Create a list by combining the two vectors.
# Lists can hold objects of different lengths and types.
my_list <- list(my_first_vector, my_second_vector)
my_list
```

Data frames are rectangular tables: rows are observations, columns are variables.

```{r}
# Create a data frame with two columns.
# Each column must have the same number of values.
my_df <- data.frame(
  author = c("Shakespeare", "Dante", "Cervantes", "Pynchon"),
  nationality = c("English", "Italian", "Spanish", "American")
)

# Print the data frame.
my_df
```

### Your Turn

Create a new data frame with authors and nationalities that are relevant to your own interests.

```{r}
# Create a new data frame.
# Replace the placeholder values with authors and nationalities of your choice.
my_new_df <- data.frame(
  author = c( "XXXXX" , "XXXXXX", "XXXXX"),
  nationality = c("XXXXX", "XXXXXX", "XXXXXX")
)

# Print your new data frame.
my_new_df
```

## Accessing Data

Indexing lets us select parts of an object. 

```{r}
# Select the first value from the logical object.
my_logical[1]

# Select the first value from the numeric object.
my_number[1]
```

A string is one value, not automatically a vector of words.

```{r}
# Print the full string.
my_string

# Extract characters 1 to 5 from the string.
# This gives characters, not words.
substr(my_string, 1, 5)
```

To access words, split the string first.

```{r}
# Split the string wherever there is a space.
# strsplit() returns a list, so [[1]] extracts the first item from that list.
words <- strsplit(my_string, " ")[[1]]

# Select the first word.
words[1]

# Select the first four words.
words[1:4]

# Select the first and fourth words.
words[c(1, 4)]

# Extract the first two letters of the first word.
substr(words[1], 1, 2)
```

Vectors, lists, and data frames each have their own selection patterns.

```{r}
# Select the first number from the numeric vector.
my_first_vector[1]
```

```{r}
# Select the first four words from the character vector.
my_second_vector[1:4]
```

```{r}
# Select the first and fourth words from the character vector.
my_second_vector[c(1, 4)]
```

```{r}
# Select the first object stored inside the list.
my_list[[1]]
```

```{r}
# Select the second object in the list, then select its first three elements.
my_list[[2]][1:3]
```

```{r}
# Select a column by name.
my_df$author
```

```{r}
# Select the first column by position.
my_df[, 1]
```

```{r}
# Select the first two values from the author column.
my_df$author[1:2]
```

```{r}
# Select the first row of the data frame.
my_df[1, ]
```

We can also select data by condition.

```{r}
# Test whether each author is Dante.
my_df$author == "Dante"
```

```{r}
# Find the row number where the author is Dante.
which(my_df$author == "Dante")
```

```{r}
# Use that row number to retrieve Dante's nationality.
my_df$nationality[which(my_df$author == "Dante")]
```

### Your Turn

Find the author with Spanish nationality.

```{r}
# Find the row where nationality is Spanish, then return the author in that row.
my_df$author[which(my_df$nationality == "Spanish")]
```

## Manipulating Values

R can perform operations on numbers and text.

```{r}
# Add 1 to every value in the vector.
my_first_vector + 1
```

```{r}
# Select the second value in the vector, then add 1 only to that value.
my_first_vector[2] + 1
```

```{r}
# Join the string with a question mark.
paste(my_string, "?")
```

```{r}
# Split the string into words; the result is a list.
strsplit(my_string, " ")
```

```{r}
# Extract the first item from the list so we have a vector of words.
strsplit(my_string, " ")[[1]]
```

```{r}
# Extract the first five characters.
substr(my_string, 1, 5)
```

### Your Turn

Extract the second occurrence of `"be"` from `my_string`.

```{r}
# Split the string into a vector of words.
words <- strsplit(my_string, " ")[[1]]

# Find all positions where the word is "be", take the second position,
# then use that position to retrieve the second "be".
words[which(words == "be")[2]]
```

## Reading and Writing Text Files

Check that your working directory is the project folder.

```{r}
# Print the current working directory.
# It should be the main project folder.
getwd()
```

Read a text file line by line.

```{r}
# Read the Austen sample file as lines of text.
# encoding = "UTF-8" helps R read special characters correctly.
my_text <- readLines("samples/austen_pride_1813.txt", encoding = "UTF-8")

# Print the first few lines.
head(my_text)
```

Collapse the lines into one string.

```{r}
# Collapse all lines into one longer string.
# The newline character "\n" keeps line breaks between the original lines.
my_text <- paste(my_text, collapse = "\n")

# Print the first 300 characters to check the result.
substr(my_text, 1, 300)
```

Write text to the console or to a file.

```{r}
# Print one string to the console.
cat("The cat is on the table")
```

```{r}
# Print two strings together.
cat("The cat is on the table", my_string)
```

```{r}
# Write a string to a new text file in the samples folder.
cat("The cat is on the table", file = "samples/Cat.txt")
```

### Your Turn

Read another sample text and split it into words.

```{r}
# Read the Melville sample text line by line.
my_text <- readLines("samples/melville_moby_1851.txt", encoding = "UTF-8")

# Collapse the lines into one string, then split at spaces.
# strsplit() returns a list.
my_words <- strsplit(paste(my_text, collapse = " "), " ")

# Extract the first element of the list, then print the first 100 words.
my_words[[1]][1:100]

```

## Packages

Packages extend what R can do. The `tidyverse` gives us many functions for data manipulation and visualisation.

```{r}
# Check whether tidyverse is installed.
# If it is not installed, install it.
if (!requireNamespace("tidyverse", quietly = TRUE)) {
  install.packages("tidyverse")
}

# Load tidyverse so that functions such as filter() and %>% are available.
library(tidyverse)
```

Compare base R and tidyverse syntax.

```{r}
# Base R version: find rows where nationality is Italian.
my_df[which(my_df$nationality == "Italian"), ]
```

```{r}
# Tidyverse version: keep rows where nationality is Italian.
my_df %>%
  filter(nationality == "Italian")
```

## Takeaway

You have used the basic building blocks we need for text analysis:

- variables
- vectors
- lists
- data frames
- indexing
- text files
- loops
- functions
- packages

<!-- Optional / extra info -->

## Loops and Conditionals

Loops repeat code.

```{r}
# Repeat the code inside the loop for every value from 1 to 10.
for (i in 1:10) {
  # Print the current value of i.
  print(i)
}
```

Conditionals let code behave differently when a condition is true.

```{r}
# Loop over the numbers 1 to 10.
for (i in 1:10) {
  # Only print i when i is exactly equal to 1.
  if (i == 1) {
    print(i)
  }
}
```

```{r}
# Loop over the numbers 1 to 10.
for (i in 1:10) {
  # If i is 4 or smaller, print the number.
  if (i <= 4) {
    print(i)
  } else {
    # Otherwise, print a message.
    print("more than four")
  }
}
```

## Functions

Functions bundle actions so we can reuse them.

```{r}
# Define a function with one input, x.
my_function <- function(x) {
  # Print "Ciao" followed by whatever value is passed to x.
  cat("Ciao", x, "\n")
}

# Run the function with one value.
my_function("Everyone!")
```

```{r}
# Define a function with two inputs.
teach_with <- function(name1, name2) {
  # Print a sentence that uses both inputs.
  cat(name1, "is teaching Text Analysis with", name2, "\n")
}

# Run the function.
teach_with("Giulia", "R")
```
