---
title: "02. Corpus Creation"
subtitle: "Text Analysis with R"
format:
  html:
    toc: true
    toc-depth: 2
    number-sections: true
execute:
  echo: true
  warning: false
  message: false
---

## Aim

This notebook shows how texts become a corpus that can be analysed in R. We import files, add metadata from filenames, split texts into sentences and tokens, and inspect the resulting data.

Corpus creation is a methodological step. The choices made here shape every later result.

```{r}
# List the packages used in this notebook.
required_packages <- c("tidyverse", "tidytext", "readtext", "readxl")

# Check which of those packages are not installed yet.
missing_packages <- required_packages[!vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)]

# Install only the packages that are missing.
if (length(missing_packages) > 0) {
  install.packages(missing_packages)
}

# Load the packages so their functions are available.
library(tidyverse)
library(tidytext)
library(readtext)
library(readxl)
```

## What We Mean by Corpus

A corpus is not just a pile of files. For this course, a corpus is a collection of texts assembled for a purpose, with enough structure to support analysis.

Important questions:

- Which texts are included?
- Which texts are missing?
- What metadata do we have?
- What does one row represent: a document, a sentence, or a word?
- What does the corpus claim to represent?

## Reading One Text File

Start with a single plain text file.

```{r}
# Print the current working directory.
# It should be the main project folder containing the samples folder.
getwd()
```

```{r}
# Read one text file as a table.
# read.delim() treats each line of the file as one row.
pride <- read.delim(
  "samples/austen_pride_1813.txt",
  fileEncoding = "UTF-8",
  header = FALSE
) %>%
  # Rename the automatically created column V1 to a clearer name.
  rename(text = V1)

# Print the first rows to check the import.
head(pride$text)
```

The file has been imported, but each row is a line/paragraph from the original file (which could contain multiple sentences). We often want sentences instead.

```{r}
# Read the text file line by line.
pride_sentences <- readLines("samples/austen_pride_1813.txt", encoding = "UTF-8") %>%
  # Collapse all lines into one long string.
  paste(collapse = " ") %>%
  # Convert the string into a one-column data frame.
  as.data.frame() %>%
  # Rename the text column.
  rename(text = ".") %>%
  # Remove repeated spaces and tidy whitespace.
  mutate(text = str_squish(text)) %>%
  # Split the text column into one sentence per row.
  unnest_sentences(
    output = "sentence",
    input = "text",
    to_lower = FALSE
  )

# Print the first sentences.
head(pride_sentences$sentence)
```

### Your Turn

Try the same process with `samples/melville_moby_1851.txt`.

```{r}
# Repeat the same workflow with the Melville sample.
melville_sentences <- readLines("samples/melville_moby_1851.txt", encoding = "UTF-8") %>%
  # Merge the original file lines into one string.
  paste(collapse = " ") %>%
  # Convert the string into a data frame.
  as.data.frame() %>%
  # Give the text column a clear name.
  rename(text = ".") %>%
  # Remove repeated spaces.
  mutate(text = str_squish(text)) %>%
  # Split the text into sentences.
  unnest_sentences(
    output = "sentence",
    input = "text",
    to_lower = FALSE
  )

# Check the first sentences.
head(melville_sentences$sentence)
```

## Reading Multiple Text Files

`readtext()` can import several files at once. The `*.txt` pattern means "all text files in this folder."

```{r}
# Read all .txt files in the samples folder.
corpus <- readtext("samples/*.txt", encoding = "UTF-8") %>%
  # Split every imported text into sentences.
  unnest_sentences(
    input = "text",
    output = "sentence",
    to_lower = FALSE,
    drop = TRUE
  ) %>%
  # Convert the result to a tibble, tidyverse's data-frame format.
  as_tibble()

# Inspect the first rows of the sentence-level corpus.
head(corpus)
```

Inspect which files were imported.

```{r}
# Show each unique document ID once.
# This lets us check exactly which files entered the corpus.
corpus %>%
  distinct(doc_id)
```

The sample folder contains a `pride_nospace.txt` file, which is useful for demonstration but not for the corpus. We remove it.

```{r}
# Remove rows from the demonstration file that contains "nospace" in its name.
corpus <- corpus %>%
  filter(!str_detect(doc_id, "nospace"))

# Check the remaining document IDs after filtering.
corpus %>%
  distinct(doc_id)
```

## Adding Metadata from Filenames

The filenames follow the pattern `author_title_year.txt`. We can turn those pieces into columns.

```{r}
# Start from the sentence-level corpus.
corpus <- corpus %>%
  # Tidy whitespace inside each sentence.
  mutate(sentence = str_squish(sentence)) %>%
  # Split the filename into metadata columns.
  separate(
    doc_id,
    into = c("author", "title", "year"),
    sep = "_",
    remove = FALSE
  ) %>%
  # Remove the .txt extension from the year and convert it to a number.
  mutate(
    year = str_remove(year, "\\.txt$"),
    year = as.numeric(year)
  )

# Inspect the corpus with the new metadata columns.
head(corpus)
```

This step matters. Metadata lets us compare texts by author, year, genre, source, or any other category we have recorded.

## Adding Sentence IDs

A sentence ID preserves order within each text.

```{r}
# Group by document so sentence numbers restart for each text.
corpus <- corpus %>%
  group_by(doc_id) %>%
  # Add a sentence number based on row order within each document.
  mutate(sentence_id = row_number()) %>%
  # Remove the grouping so later operations use the whole data frame.
  ungroup()

# Check the sentence IDs.
head(corpus)
```

## Tokenising the Corpus

Tokenisation splits text into units, usually words.

```{r}
# Split each sentence into tokens, usually words.
corpus_token <- corpus %>%
  unnest_tokens(
    input = "sentence",
    output = "token",
    to_lower = FALSE,
    drop = FALSE
  )

# Inspect the first ten token rows.
head(corpus_token, 10)
```

Add a token ID to preserve word order inside each sentence.

```{r}
# Group by document and sentence so token numbers restart in each sentence.
corpus_token <- corpus_token %>%
  group_by(doc_id, sentence_id) %>%
  # Add a token number based on word order within the sentence.
  mutate(token_id = row_number()) %>%
  # Remove grouping for later analysis.
  ungroup()

# Check token order.
head(corpus_token, 10)
```

## Reading Spreadsheet Data

Texts and metadata may also arrive in spreadsheets.

```{r}
# Read an Excel file from the samples folder.
pride_excel <- read_excel("samples/pride.xlsx")

# Inspect the first rows.
head(pride_excel)
```

`readtext()` can also import spreadsheet files from a folder.

```{r}
# Read all Excel files in the samples folder.
corpus_source <- readtext("samples/*.xlsx")

# Inspect the imported spreadsheet-based corpus.
head(corpus_source)
```

## Checkpoint

By this point we have created:

- a sentence-level corpus
- a token-level corpus
- metadata columns from filenames
- sentence and token IDs

Before moving to analysis, ask:

- Is the corpus suitable for the question?
- What does it represent?
- What has been excluded?
- What metadata would make the analysis stronger?
