---
title: "02. Corpus Creation"
subtitle: "Text Analysis with R"
author: "Dr Giulia Grisot"
format:
  html:
    toc: true
    toc-depth: 2
    number-sections: true
execute:
  echo: true
  warning: false
  message: false
---

> Author: Dr Giulia Grisot. This script is distributed under the MIT licence.

## Aim

This notebook shows how texts become a corpus that can be analysed in R.

To keep the workflow simple, we use **one import strategy only**: the `readtext` package. We use it first for one text file, then for a folder of text files. Later scripts build on the same kind of corpus object.

Corpus creation is a methodological step. The choices made here shape every later result.

```{r}
# List the packages used in this notebook.
required_packages <- c("tidyverse", "tidytext", "readtext")

# Check which packages are not installed yet.
missing_packages <- required_packages[!vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)]

# Install only the packages that are missing.
if (length(missing_packages) > 0) {
  install.packages(missing_packages)
}

# Load the packages so their functions are available.
library(tidyverse)
library(tidytext)
library(readtext)
```

## What We Mean by Corpus

A corpus is not just a pile of files. For this course, a corpus is a collection of texts assembled for a purpose, with enough structure to support analysis.

Important questions:

- Which texts are included?
- Which texts are missing?
- What metadata do we have?
- What does one row represent: a document, a sentence, or a word?
- What does the corpus claim to represent?

## Check the Project Folder

Before reading files, check that R is working from the project folder.

```{r}
# Print the current working directory.
# It should be the main project folder containing the samples folder.
getwd()
```

## Read One Text File with `readtext`

Start with one plain text file.

```{r}
# Read one text file with readtext().
# The result has one row per file and includes a doc_id column.
pride <- readtext("samples/austen_pride_1813.txt", encoding = "UTF-8")

# Inspect the imported document.
pride
```

`readtext()` gives us a document-level object: one row for the file, with the full text in a `text` column.

To analyse the text, we often want smaller units, such as sentences.

```{r}
# Start from the document-level object.
pride_sentences <- pride %>%
  # Split the text column into one sentence per row.
  unnest_sentences(
    output = "sentence",
    input = "text",
    to_lower = FALSE
  ) %>%
  # Remove repeated spaces inside sentences.
  mutate(sentence = str_squish(sentence))

# Print the first sentences.
head(pride_sentences$sentence)
```

### Your Turn

Use the same `readtext()` workflow for `samples/melville_moby_1851.txt`. Replace the 'XXXX" with the correct file path. 

```{r}
# Read one Melville sample file with readtext().
melville <- readtext("XXXXX", encoding = "UTF-8")

# Split the Melville text into sentences.
melville_sentences <- melville %>%
  unnest_sentences(
    output = "sentence",
    input = "text",
    to_lower = FALSE
  ) %>%
  mutate(sentence = str_squish(sentence))

# Check the first sentences.
head(melville_sentences$sentence)
```

## Read Multiple Text Files with `readtext`

The same package can read many files at once. The pattern `samples/*.txt` means "all `.txt` files in the samples folder."

```{r}
# Read all .txt files in the samples folder.
# Each file becomes one row.
corpus_docs_small <- readtext("samples/*.txt", encoding = "UTF-8") %>%
  # Convert to a tibble for tidyverse workflows.
  as_tibble()

# Inspect the document-level corpus.
head(corpus_docs_small)
```

The sample folder contains a `pride_nospace.txt` file, which is useful for demonstration but not for the corpus. Also the `Cat.txt` document we created as a test is unnecessary here. We remove them.

```{r}
# Remove rows from the demonstration file that contains "nospace" in its name.
corpus_docs_small <- corpus_docs_small %>%
  filter(!str_detect(doc_id, "nospace")) %>%
  filter(!str_detect(doc_id, "Cat"))

# Check the remaining document IDs after filtering.
corpus_docs_small %>%
  distinct(doc_id)
```

## Add Metadata from Filenames

The filenames follow the pattern `author_title_year.txt`. We can turn those pieces into columns.

```{r}
# Start from the document-level corpus.
corpus_docs_small <- corpus_docs_small %>%
  # Split the filename into metadata columns.
  separate(
    doc_id,
    into = c("author", "title", "year"),
    sep = "_",
    remove = FALSE
  ) %>%
  # Remove the .txt extension from year and convert it to a number.
  mutate(
    year = str_remove(year, "\\.txt$"),
    year = as.numeric(year)
  )

# Inspect the document-level corpus with metadata.
corpus_docs_small %>%
  select(doc_id, author, title, year) %>%
  head()
```

This step matters. Metadata lets us compare texts by author, year, genre, source, or any other category we have recorded.

## Create a Sentence-Level Corpus

Now we split every document into sentences.

```{r}
# Start from the document-level corpus.
corpus_sentence_small <- corpus_docs_small %>%
  # Split each document's text into one sentence per row.
  unnest_sentences(
    input = "text",
    output = "sentence",
    to_lower = FALSE,
    drop = TRUE
  ) %>%
  # Remove repeated spaces inside sentences.
  mutate(sentence = str_squish(sentence)) %>%
  # Group by document so sentence numbers restart for each text.
  group_by(doc_id) %>%
  # Add a sentence number based on row order within each document.
  mutate(sentence_id = row_number()) %>%
  # Remove the grouping so later operations use the whole data frame.
  ungroup()

# Inspect the sentence-level corpus.
head(corpus_sentence_small)
```

## Check back for errors

Sentence splitting is never completely neutral. A common problem is that abbreviations such as `Mr.`, `Mrs.`, or `Dr.` contain a full stop, so R may treat them as complete sentences.

First, let us check whether this happened.

```{r}
# Look for very short sentences that are probably abbreviations.
# The pattern means: start of sentence, one common title, full stop, end of sentence.
corpus_sentence_small %>%
  filter(str_detect(sentence, "^(Mr|Mrs|Ms|Miss|Dr|Prof|Rev|Hon|Capt|Col|Gen|Sgt|Jr|Sr|St|Mme|Mlle)\\.$")) %>%
  count(sentence, sort = TRUE)
```

It is also useful to inspect very short sentences more generally. Some will be real sentences, but others may reveal sentence-splitting errors.

```{r}
# Count sentences with one or two visible word-like units.
# This is a quick diagnostic, not a perfect error detector.
corpus_sentence_small %>%
  filter(str_count(sentence, "\\S+") <= 2) %>%
  count(sentence, sort = TRUE) %>%
  head(20)
```

One pragmatic fix is to temporarily protect the full stops in common titles before sentence splitting. After splitting, we put the full stops back.

This is a partial solution: it handles common abbreviations, but it will not solve every possible sentence-splitting problem. It may also make the wrong decision in rare cases where an abbreviation genuinely appears at the end of a sentence. The important thing is that we have noticed the issue, made a decision, and can explain what we changed.

```{r}
# Define a helper function to protect full stops in common title abbreviations.
# For example, "Mr." becomes "Mr<ABBR_DOT>" before sentence splitting.
protect_title_dots <- function(text) {
  str_replace_all(
    text,
    "\\b(Mr|Mrs|Ms|Miss|Dr|Prof|Rev|Hon|Capt|Col|Gen|Sgt|Jr|Sr|St|Mme|Mlle)\\.",
    "\\1<ABBR_DOT>"
  )
}

# Define a second helper function to restore the protected full stops.
# For example, "Mr<ABBR_DOT>" becomes "Mr." again.
restore_title_dots <- function(text) {
  str_replace_all(text, "<ABBR_DOT>", ".")
}
```

Now we recreate the sentence-level corpus with this small preprocessing step.

```{r}
# Start again from the document-level corpus.
corpus_sentence_small <- corpus_docs_small %>%
  # Protect dots in common title abbreviations before splitting sentences.
  mutate(text = protect_title_dots(text)) %>%
  # Split each document's text into one sentence per row.
  unnest_sentences(
    input = "text",
    output = "sentence",
    to_lower = FALSE,
    drop = TRUE
  ) %>%
  # Restore the dots in abbreviations after sentence splitting.
  mutate(sentence = restore_title_dots(sentence)) %>%
  # Remove repeated spaces inside sentences.
  mutate(sentence = str_squish(sentence)) %>%
  # Group by document so sentence numbers restart for each text.
  group_by(doc_id) %>%
  # Recreate sentence IDs after fixing the sentence split.
  mutate(sentence_id = row_number()) %>%
  # Remove grouping for later analysis.
  ungroup()

# Inspect the first rows of the corrected sentence-level corpus.
head(corpus_sentence_small)
```

Check whether the abbreviation-only sentences are still present.

```{r}
# Run the same diagnostic again after the fix.
corpus_sentence_small %>%
  filter(str_detect(sentence, "^(Mr|Mrs|Ms|Miss|Dr|Prof|Rev|Hon|Capt|Col|Gen|Sgt|Jr|Sr|St|Mme|Mlle)\\.$")) %>%
  count(sentence, sort = TRUE)
```


## Create a Token-Level Corpus

Tokenisation splits text into units, usually words.

```{r}
# Start from the sentence-level corpus.
corpus_token_small <- corpus_sentence_small %>%
  # Split each sentence into tokens.
  unnest_tokens(
    input = "sentence",
    output = "token",
    to_lower = FALSE,
    drop = FALSE
  )

# Inspect the first ten token rows.
head(corpus_token_small, 10)
```

Add a token ID to preserve word order inside each sentence.

```{r}
# Group by document and sentence so token numbers restart in each sentence.
corpus_token_small <- corpus_token_small %>%
  group_by(doc_id, sentence_id) %>%
  # Add a token number based on word order within the sentence.
  mutate(token_id = row_number()) %>%
  # Remove grouping for later analysis.
  ungroup()

# Check token order.
head(corpus_token_small, 10)
```

## Checkpoint

By this point we have created:

- a document-level corpus
- a sentence-level corpus
- a token-level corpus
- metadata columns from filenames
- sentence and token IDs

Before moving to analysis, ask:

- Is the corpus suitable for the question?
- What does it represent?
- What has been excluded?
- What metadata would make the analysis stronger?
