---
title: "03. Corpus Analysis"
subtitle: "Text Analysis with R"
author: "Dr Giulia Grisot"
format:
  html:
    toc: true
    toc-depth: 2
    number-sections: true
execute:
  echo: true
  warning: false
  message: false
---

> Author: Dr Giulia Grisot. This script is distributed under the MIT licence.

## Aim

This notebook introduces core corpus analysis methods:

- inspecting corpus structure
- counting words
- removing stopwords
- comparing authors
- finding n-grams
- using concordances / KWIC
- tracing word frequencies over time
- finding co-occurrence in sentences

The notebook uses precomputed `.RData` objects so that students can work quickly in RStudio or Posit Cloud.

```{r}
# List the packages used in this notebook.
required_packages <- c("tidyverse", "tidytext", "readtext", "quanteda")

# Identify packages that are not installed.
missing_packages <- required_packages[!vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)]

# Install only missing packages.
if (length(missing_packages) > 0) {
  install.packages(missing_packages)
}

# Load the packages.
library(tidyverse)
library(tidytext)
library(readtext)
library(quanteda)
```

## Load the Prepared Corpus

The repository includes three prepared corpus objects:

- `corpus_token_sample`: one row per token
- `corpus_sentence`: one row per sentence
- `corpus_docs`: one row per document

```{r}
# Load the token-level corpus: one row per token.
load("corpus_token_sample.RData")

# Load the sentence-level corpus: one row per sentence.
load("corpus_sentence.RData")

# Load the document-level corpus: one row per full text.
load("corpus_docs.RData")
```

## Inspect the Data

Always inspect a corpus before analysing it.

```{r}
# Show the structure of the token-level corpus.
# This tells us the column names and data types.
glimpse(corpus_token_sample)
```

```{r}
# Print the first rows so we can see what the data looks like.
head(corpus_token_sample)
```

Basic corpus size:

```{r}
# Count the total number of token rows.
nrow(corpus_token_sample)

# Count the number of unique word forms.
length(unique(corpus_token_sample$token))

# Count the number of unique authors.
length(unique(corpus_token_sample$author))
```

For this session, keep only the columns we need most often.

```{r}
# Keep only the columns needed for the exercises below.
corpus_token_sample <- corpus_token_sample %>%
  select(doc_id, sentence, token, sentence_id, title, author, year)
```

## Word Frequencies

The simplest question is: which words occur most often?

```{r}
# Count how often each token appears.
# sort = TRUE puts the most frequent tokens at the top.
token_freq <- corpus_token_sample %>%
  count(token, sort = TRUE)

# Print the 20 most frequent tokens.
head(token_freq, 20)
```

Many of the most frequent words are function words. Stopword lists help remove common words, but they are always a methodological choice.

```{r}
# Read the stopword list from a text file.
stopwords_en <- readtext("jockers_stopwords.txt") %>%
  # Split the stopword text into one word per row.
  unnest_tokens(input = "text", output = "stopword") %>%
  # Create a lowercase token column to match against the corpus.
  mutate(token = tolower(stopword)) %>%
  # Keep only the column needed for joining/filtering.
  select(token)
```

```{r}
# Start from the token frequency table.
token_freq %>%
  # Convert tokens to lowercase so "The" and "the" are treated alike.
  mutate(token = tolower(token)) %>%
  # Remove tokens that appear in the stopword list.
  anti_join(stopwords_en, by = "token") %>%
  # Keep the 20 most frequent remaining tokens.
  slice_max(n, n = 20) %>%
  # Put token frequency on a bar chart.
  ggplot(aes(x = reorder(token, n), y = n)) +
  geom_col(fill = "steelblue") +
  # Flip the chart so token labels are easier to read.
  coord_flip() +
  labs(
    title = "Top 20 Tokens After Stopword Removal",
    x = "Token",
    y = "Frequency"
  )
```

### Your Turn

How many times does `"love"` appear?

```{r}
# Keep only rows where the token is exactly "love".
corpus_token_sample %>%
  filter(token == "love") %>%
  # Count the remaining rows.
  nrow()
```

Compare top words for different authors.

```{r}
# Define a reusable function for finding frequent words by author.
top_tokens_by_author <- function(author_pattern, n_terms = 10) {
  # Start from the token-level corpus.
  corpus_token_sample %>%
    # Lowercase tokens for consistency.
    mutate(token = tolower(token)) %>%
    # Remove stopwords.
    anti_join(stopwords_en, by = "token") %>%
    # Keep authors whose names match the pattern.
    filter(str_detect(author, author_pattern)) %>%
    # Count token frequency.
    count(token, sort = TRUE) %>%
    # Keep the top results.
    slice_head(n = n_terms)
}

# Run the function for Woolf.
top_tokens_by_author("Woolf")
```

```{r}
# Run the same function for Dickens.
top_tokens_by_author("Dickens")
```

Try another author:

```{r}
# Try a pattern that matches several author names.
top_tokens_by_author("Austen|Bronte|Eliot")
```

## N-Grams and Collocations

N-grams are sequences of words. They can help identify repeated phrases.

```{r}
# Start from the document-level corpus.
ngrams <- corpus_docs %>%
  # Split each document into 4-word sequences.
  unnest_ngrams(input = text, output = ngram, n = 4) %>%
  # Count each repeated 4-word sequence.
  count(ngram, sort = TRUE) %>%
  # Keep only phrases that appear more than 10 times.
  filter(n > 10)

# Print the most frequent n-grams.
head(ngrams, 20)
```

Compare repeated phrases for selected authors.

```{r}
# Start from the document-level corpus.
corpus_docs %>%
  # Keep only selected authors for comparison.
  filter(str_detect(author, "Woolf|Dickens|Brontë|Carroll")) %>%
  # Calculate n-grams separately for each author.
  group_by(author) %>%
  # Split each author's texts into 4-word sequences.
  unnest_ngrams(input = text, output = ngram, n = 4) %>%
  # Count repeated n-grams.
  count(ngram, sort = TRUE) %>%
  # Keep the first three rows per author group.
  slice_head(n = 3)
```

### Your Turn

Change `n = 4` to `n = 2` or `n = 3`. What different kinds of phrases become visible?

## Concordance / KWIC

A concordance shows a search term in context. KWIC means "keyword in context."

```{r}
# Convert the document-level data frame into a quanteda corpus.
corpus_q <- quanteda::corpus(corpus_docs, text_field = "text")

# Tokenise the quanteda corpus.
corpus_q_tokens <- tokens(
  corpus_q,
  # Remove punctuation marks.
  remove_punct = TRUE,
  # Remove symbols.
  remove_symbols = TRUE
) %>%
  # Lowercase all tokens.
  tokens_tolower()
```

Search for one word.

```{r}
# Search for the word "power" with three words of context on each side.
kwic(corpus_q_tokens, pattern = "power", window = 3) %>%
  # Convert the result to a regular data frame for display.
  as.data.frame() %>%
  # Show the first ten results.
  head(10)
```

Search for word forms with a wildcard.

```{r}
# Search for words beginning with "beaut", such as beauty or beautiful.
kwic(corpus_q_tokens, pattern = "beaut*", window = 3) %>%
  as.data.frame() %>%
  head(10)
```

Search for a phrase.

```{r}
# Search for the phrase "married man".
# phrase() tells quanteda to treat the words as a phrase, not separate terms.
kwic(corpus_q_tokens, pattern = phrase("married man"), window = 3) %>%
  as.data.frame() %>%
  head(10)
```

### Your Turn

Try another word, phrase, or wildcard. Change `window` to see more or less context.

## Frequency Over Time

If metadata includes dates, we can ask how word frequency changes over time.

```{r}
# Make sure year is numeric so it can be plotted on a continuous x-axis.
corpus_token_sample <- corpus_token_sample %>%
  mutate(year = as.numeric(year))
```

```{r}
# Start from the token-level corpus.
corpus_token_sample %>%
  # Keep only rows where the token is "war".
  filter(token == "war") %>%
  # Count how many times it appears in each year.
  count(year) %>%
  # Plot year on the x-axis and frequency on the y-axis.
  ggplot(aes(x = year, y = n)) +
  geom_line(color = "darkred") +
  labs(
    title = "Frequency of 'war' Over Time",
    x = "Year",
    y = "Frequency"
  )
```

```{r}
# Repeat the same workflow for the word "myself".
corpus_token_sample %>%
  filter(token == "myself") %>%
  count(year) %>%
  ggplot(aes(x = year, y = n)) +
  geom_line(color = "steelblue") +
  labs(
    title = "Frequency of 'myself' Over Time",
    x = "Year",
    y = "Frequency"
  )
```

### Your Turn

Plot a different word, such as `"peace"`, `"love"`, or a term from your own research.

## Co-Occurrence Within Sentences

Co-occurrence asks whether two terms appear in the same context unit. Here the context unit is the sentence.

```{r}
# Start from the sentence-level corpus.
co_occurrence <- corpus_sentence %>%
  # Keep sentences containing "war".
  filter(
    str_detect(sentence, regex("war", ignore_case = TRUE)),
    # Also keep only those that contain "peace".
    str_detect(sentence, regex("peace", ignore_case = TRUE))
  ) %>%
  # Keep the identifying columns and the sentence text.
  select(doc_id, sentence_id, sentence)

# Display the first ten matching sentences.
head(co_occurrence, 10)
```

### Your Turn

Try another pair of words. What does the sentence context reveal that a frequency count would not?

## Takeaway

Corpus analysis helps us move between scale and context:

- counts show patterns
- stopwords and metadata shape the result
- n-grams and collocations suggest repeated associations
- KWIC returns us to language in context
- time-based plots depend on metadata
- co-occurrence can identify sentences for close reading
