---
title: "04. Sentiment Analysis"
subtitle: "Text Analysis with R"
author: "Dr Giulia Grisot"
format:
  html:
    toc: true
    toc-depth: 2
    number-sections: true
execute:
  echo: true
  warning: false
  message: false
---

> Author: Dr Giulia Grisot. This script is distributed under the MIT licence.

## Aim

This notebook introduces lexicon-based sentiment analysis. We will compare sentiment methods, join a sentiment lexicon to corpus tokens, visualise sentiment across texts, inspect sentiment in sentences, and consider negation.

Sentiment analysis does not directly measure meaning, emotion, or reader response. It is a method for producing interpretable patterns that need to be checked against text and context.

```{r}
# List the packages needed for sentiment analysis.
required_packages <- c("tidyverse", "syuzhet", "tidytext", "plotly", "sentimentr")

# Find which packages are missing from this R installation.
missing_packages <- required_packages[!vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)]

# Install only the missing packages.
if (length(missing_packages) > 0) {
  install.packages(missing_packages)
}

# Load the core packages for this notebook.
library(tidyverse)
library(syuzhet)
library(tidytext)
```

## Load the Corpus

```{r}
# Load the token-level corpus used for lexicon matching.
load("corpus_token_sample.RData")

# Load the sentence-level corpus used for sentence plots.
load("corpus_sentence.RData")
```

Check what documents are present.

```{r}
# Show each author/document/year combination once.
corpus_sentence %>%
  distinct(author, doc_id, year)
```

```{r}
# Do the same check in the token-level corpus.
corpus_token_sample %>%
  distinct(author, doc_id, year)
```

How many authors are represented?

```{r}
# Select unique author names, then count them.
corpus_token_sample %>%
  distinct(author) %>%
  nrow()
```

### Your Turn

Which years are present?

```{r}
# List the unique years in order.
corpus_token_sample$year %>%
  summary()
```

```{r}
# Count how many unique years are present.
corpus_token_sample %>%
  distinct(year) %>%
  nrow()
```

## We can also visualise this

```{r}
corpus_token_sample %>%
  select(doc_id, year) %>%
  distinct() %>%
  ggplot(aes(year)) +
  geom_histogram()
```


## Sentiment Lexicons

A sentiment lexicon is a list of words with associated sentiment values or categories.

```{r}
# Load the NRC sentiment dictionary through syuzhet.
syuzhet_nrc <- get_sentiment_dictionary("nrc", language = "english")

# Load the syuzhet sentiment dictionary.
syuzhet_en <- get_sentiment_dictionary("syuzhet", language = "english")

# Inspect the first rows of the lexicon.
head(syuzhet_en)
```

Different lexicons can produce different results.

```{r}
# Score the same sentence with different lexicon methods.
# Different methods can produce different scores.
get_sentiment("I like apples and pears", language = "english", method = "syuzhet")
get_sentiment("I like apples and pears", language = "english", method = "bing")
get_sentiment("I like apples and pears", language = "english", method = "nrc")
get_sentiment("I like apples and pears", language = "english", method = "afinn")
```

```{r}
# Score a sentence with darker affective language.
get_sentiment("She was walking alone in a dark and eerie night", language = "english", method = "syuzhet")
```

## Join the Lexicon to the Corpus

Joining a lexicon to a large token table can take time. We do it once and reuse the result.

```{r}
# Start with the token-level corpus.
corpus_sentiment_words <- corpus_token_sample %>%
  # Create a lowercase word column so it matches the lexicon format.
  mutate(word = tolower(token)) %>%
  # Join sentiment values onto matching words.
  # Words not found in the lexicon receive NA for value.
  left_join(syuzhet_en, by = "word", relationship = "many-to-many")

# Inspect the joined data.
head(corpus_sentiment_words)
```

## Average Sentiment by Text

```{r}
# Group words by author and title so we can summarise each text.
sentiment_by_text <- corpus_sentiment_words %>%
  group_by(author, title) %>%
  # Average the sentiment values, ignoring words with no lexicon value.
  summarise(sentiment_value = mean(value, na.rm = TRUE), .groups = "drop")

# Print the resulting text-level sentiment table.
sentiment_by_text
```

```{r}
# Plot the text-level sentiment table.
sentiment_by_text %>%
  # Put sentiment on the x-axis and order titles by their score.
  ggplot(aes(x = sentiment_value, y = reorder(title, sentiment_value))) +
  # Draw one bar per title.
  geom_col(fill = "steelblue") +
  # Flip coordinates so long titles are easier to read.
  coord_flip() +
  labs(
    title = "Average Sentiment by Text",
    x = "Average sentiment value",
    y = "Title"
  )
```

Optional interactive version:

```{r}
# Load plotly for interactive plots.
library(plotly)

# Build a ggplot object first.
p <- sentiment_by_text %>%
  ggplot(aes(
    x = sentiment_value,
    y = reorder(title, sentiment_value),
    # This text will appear in the interactive tooltip.
    text = paste(author, title, sep = ": ")
  )) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(x = "Average sentiment value", y = "Title")

# Convert the ggplot into an interactive plotly plot.
ggplotly(p, tooltip = "text")
```

## Sentiment by Year

```{r}
# Group by title, author, and year so each point represents one text.
sentiment_by_year <- corpus_sentiment_words %>%
  group_by(title, author, year) %>%
  # Average sentiment values within each text.
  summarise(sentiment_value = mean(value, na.rm = TRUE), .groups = "drop")

# Plot publication year against average sentiment.
sentiment_by_year %>%
  ggplot(aes(x = year, y = sentiment_value)) +
  geom_point(size = 2) +
  geom_smooth(se = FALSE) +
  labs(
    title = "Average Sentiment by Publication Year",
    x = "Year",
    y = "Average sentiment value"
  )
```

## Positive and Negative Sentences

Calculate sentiment values by sentence.

```{r}
# Group by document and sentence so each row becomes one sentence score.
sentence_sentiment <- corpus_sentiment_words %>%
  group_by(doc_id, sentence_id, title, author) %>%
  # Add together sentiment values for all matched words in a sentence.
  summarise(value = sum(value, na.rm = TRUE), .groups = "drop") %>%
  # Put the most positive sentences at the top.
  arrange(desc(value))

# Show the most positive sentences.
head(sentence_sentiment, 10)
```

```{r}
# Show the most negative sentences.
tail(sentence_sentiment, 10)
```

## Inspect One Sentence

Choose one sentence and show which words contributed to the score.

```{r}
# Select one sentence from one document.
test_sentence <- corpus_token_sample %>%
  filter(sentence_id == 212, doc_id == "ENG18670_Ouida") %>%
  # Lowercase tokens so they match the lexicon.
  mutate(word = tolower(token)) %>%
  # Add sentiment values for matched words.
  left_join(syuzhet_en, by = "word", relationship = "many-to-many")

# Show each token and its sentiment value.
test_sentence %>%
  select(token, value) %>%
  head(20)
```

```{r}
# Start with the selected sentence.
test_sentence %>%
  mutate(
    # Break the sentence into lines of ten tokens for plotting.
    line_number = rep(1:ceiling(n() / 10), each = 10, length.out = n()),
    # Give each token a position from 1 to 10 within each plotted line.
    token_position = rep(1:10, length.out = n())
  ) %>%
  # Plot token position horizontally and line number vertically.
  ggplot(aes(x = token_position, y = -line_number, fill = value)) +
  # Draw one tile behind each token.
  geom_tile() +
  # Print the token text on top of each tile.
  geom_text(aes(label = token), size = 3) +
  # Use colour to show sentiment value; grey means no sentiment match.
  scale_fill_viridis_c(na.value = "grey90") +
  theme_minimal() +
  theme(
    axis.title = element_blank(),
    legend.position = "none"
  ) +
  labs(
    title = "Sentiment Values Within One Sentence",
    subtitle = paste0(test_sentence$title[1], " by ", test_sentence$author[1])
  )
```

### Your Turn

In this exercise, you will choose one short sentence with a clear sentiment score, inspect which words produced that score, and then make your own heatmap.

The first chunk creates a beginner-friendly list of candidate sentences. It keeps sentences that:

- are not too long
- have at least one sentiment word
- have a strong positive or negative score

```{r}
# Step 1: count how many tokens are in each sentence.
sentence_lengths <- corpus_token_sample %>%
  count(doc_id, sentence_id, name = "sentence_length")

# Step 2: create a short list of sentences that are easier to inspect.
candidate_sentences <- sentence_sentiment %>%
  # Add the sentence length information.
  left_join(sentence_lengths, by = c("doc_id", "sentence_id")) %>%
  # Keep sentences of 35 tokens or fewer so the heatmap is readable.
  filter(sentence_length <= 35) %>%
  # Keep only sentences with some sentiment.
  filter(value != 0) %>%
  # Make a helper column for the strength of the sentiment score.
  mutate(abs_value = abs(value)) %>%
  # Put the strongest sentiment examples first.
  arrange(desc(abs_value)) %>%
  # Keep the first ten examples.
  slice_head(n = 10) %>%
  # Add a simple row number so we can choose one sentence.
  mutate(choice_number = row_number()) %>%
  # Show the most useful columns.
  select(choice_number, title, author, doc_id, sentence_id, sentence_length, value)

# Print the candidate sentences.
candidate_sentences
```

Choose one row from `candidate_sentences`. To choose a different sentence, change the number assigned to `my_choice`.

```{r}
# Step 3: choose which sentence you want to inspect.
# Pick a number from the choice_number column above.
my_choice <- 1

# Keep only the row you chose.
chosen_sentence <- candidate_sentences %>%
  filter(choice_number == my_choice)

# Print the sentence information.
chosen_sentence
```

```{r}
# Step 4: retrieve the actual words from the chosen sentence.
chosen_sentence_words <- corpus_token_sample %>%
  # Keep only the document selected above.
  filter(doc_id == chosen_sentence$doc_id[1]) %>%
  # Keep only the sentence selected above.
  filter(sentence_id == chosen_sentence$sentence_id[1]) %>%
  # Keep the words in their original sentence order.
  arrange(token_id) %>%
  # Lowercase tokens so they match the sentiment lexicon.
  mutate(word = tolower(token)) %>%
  # Add sentiment values from the lexicon.
  left_join(syuzhet_en, by = "word", relationship = "many-to-many")

# Print the full sentence first.
chosen_sentence_words$sentence[1]
```

```{r}
# Step 5: show each token and its sentiment value.
# NA means that the word was not found in the sentiment lexicon.
chosen_sentence_words %>%
  select(token, value)
```

```{r}
# Step 6: make a heatmap of the chosen sentence.
chosen_sentence_words %>%
  mutate(
    # Break the sentence into lines of ten tokens for plotting.
    line_number = rep(1:ceiling(n() / 10), each = 10, length.out = n()),
    # Give each token a position from 1 to 10 within each plotted line.
    token_position = rep(1:10, length.out = n())
  ) %>%
  # Plot token position horizontally and line number vertically.
  ggplot(aes(x = token_position, y = -line_number, fill = value)) +
  # Draw one tile behind each token.
  geom_tile() +
  # Print the token text on top of each tile.
  geom_text(aes(label = token), size = 3) +
  # Use colour to show sentiment value; grey means no sentiment match.
  scale_fill_viridis_c(na.value = "grey90") +
  theme_minimal() +
  theme(
    axis.title = element_blank(),
    legend.position = "none"
  ) +
  labs(
    title = "Sentiment Values Within Your Chosen Sentence",
    subtitle = paste0(chosen_sentence_words$title[1], " by ", chosen_sentence_words$author[1])
  )
```

Questions to ask:

- Which words have positive or negative sentiment values?
- Are any important words missing from the lexicon?
- Does the heatmap match your own reading of the sentence?


## Sentiment Over Narrative Time

We can also plot sentence-level sentiment across one document.

```{r}
# Start from the sentence-level corpus.
corpus_sentence %>%
  # Keep one document for a narrative-time example.
  filter(doc_id == "ENG18471_Bronte") %>%
  # Calculate a syuzhet sentiment score for each sentence.
  mutate(sentiment_syuzhet = get_sentiment(sentence)) %>%
  # Plot sentiment against sentence order.
  ggplot(aes(x = sentence_id, y = sentiment_syuzhet)) +
  geom_point(alpha = 0.5) +
  geom_smooth(se = FALSE) +
  labs(
    title = "Sentence Sentiment Across One Text",
    x = "Sentence ID",
    y = "Sentiment value"
  )
```

## Negation with `sentimentr`

Lexicon approaches can struggle with negation. The `sentimentr` package handles negation more explicitly.

```{r}
# Load sentimentr for negation-sensitive sentiment scoring.
library(sentimentr)

# Score a positive sentence.
sentiment("I like apples and pears")

# Score a similar sentence with negation.
sentiment("I don't like apples and pears")
```

The full `sentimentr` calculation has been precomputed for the class.

```{r}
# Load the precomputed sentimentr results.
load("sentimentr_sent_values.RData")

# Inspect the first rows.
head(sentimentr_sent_values)
```

```{r}
# Start from the precomputed sentence-level sentiment table.
sentimentr_sent_values %>%
  # Keep one document for the example plot.
  filter(doc_id == "ENG18471_Bronte") %>%
  # Plot sentence order against sentiment value.
  ggplot(aes(x = sentence_id, y = sentiment_value)) +
  geom_point(alpha = 0.5) +
  geom_smooth(se = FALSE) +
  labs(
    title = "Sentence Sentiment with sentimentr",
    subtitle = "Negation-sensitive polarity scores",
    x = "Sentence ID",
    y = "Sentiment value"
  )
```

Compare average sentiment by document using `sentimentr`.

```{r}
# Start from sentence-level sentimentr values.
sentimentr_sent_values %>%
  # Group by document.
  group_by(doc_id) %>%
  # Average sentence scores into one document-level score.
  summarise(sentiment_value = mean(sentiment_value, na.rm = TRUE), .groups = "drop") %>%
  # Add document metadata back in.
  left_join(
    corpus_sentence %>%
      distinct(doc_id, author, title, year),
    by = "doc_id"
  ) %>%
  # Plot one bar per title.
  ggplot(aes(x = sentiment_value, y = reorder(title, sentiment_value))) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(
    title = "Average Document Sentiment with sentimentr",
    x = "Average sentiment value",
    y = "Title"
  )
```

## Guided Exercises

Use these templates to adapt the analysis. The values in quotation marks are the parts to change first.

### Exercise 1: Compare Lexicons on Your Own Sentence

```{r}
# Write a sentence you want to test.
my_sentence <- "I was happy, but the ending made me uneasy."

# Compare several lexicon methods on the same sentence.
tibble(
  method = c("syuzhet", "bing", "nrc", "afinn"),
  sentiment = c(
    get_sentiment(my_sentence, method = "syuzhet"),
    get_sentiment(my_sentence, method = "bing"),
    get_sentiment(my_sentence, method = "nrc"),
    get_sentiment(my_sentence, method = "afinn")
  )
)
```

Questions to ask:

- Do the methods agree?
- Which words do you think are driving the result?
- Does the score match your reading of the sentence?

### Exercise 2: Focus on One Author

```{r}
# Choose one or more author patterns to investigate.
# Try one name, such as "Woolf", or several names, such as c("Woolf", "Dickens").
target_author <- c("Woolf", "Dickens", "Trollope")

# Combine the selected names into one search pattern.
# For example, c("Woolf", "Dickens") becomes "Woolf|Dickens".
author_pattern <- str_c(target_author, collapse = "|")

# Create a readable label for plot titles.
author_label <- str_c(target_author, collapse = ", ")

# Calculate average sentiment for texts by the selected author or authors.
author_sentiment <- corpus_sentiment_words %>%
  # Keep rows where the author column contains any selected name.
  filter(str_detect(author, regex(author_pattern, ignore_case = TRUE))) %>%
  group_by(author, title, year) %>%
  summarise(sentiment_value = mean(value, na.rm = TRUE), .groups = "drop")

# Print the result.
author_sentiment
```

```{r}
# Plot the selected author or authors.
author_sentiment %>%
  ggplot(aes(x = year, y = sentiment_value, colour = author, label = title)) +
  geom_point(size = 3) +
  geom_text(vjust = -0.8, size = 3, show.legend = FALSE) +
  labs(
    title = paste("Average sentiment for", author_label),
    x = "Year",
    y = "Average sentiment value",
    colour = "Author"
  )
```

Questions to ask:

- Are there differences between texts?
- Are the differences large enough to interpret?
- Which text would you inspect more closely?

### Exercise 3: Compare Two Documents

```{r}
# Choose two document IDs from the corpus.
# Use distinct(author, title, doc_id) if you need to look them up.
doc_a <- "ENG18471_Bronte"
doc_b <- "ENG19192_Woolf"

# Calculate sentence-level sentiment for both documents.
two_docs_sentiment <- sentimentr_sent_values %>%
  filter(doc_id %in% c(doc_a, doc_b))

# Add title and author metadata.
two_docs_sentiment <- two_docs_sentiment %>%
  left_join(
    corpus_sentence %>%
      distinct(doc_id, author, title),
    by = "doc_id"
  )

# Inspect the first rows.
head(two_docs_sentiment)
```

```{r}
# Plot sentiment over sentence order for the two selected documents.
two_docs_sentiment %>%
  ggplot(aes(x = sentence_id, y = sentiment_value, colour = title)) +
  geom_point(alpha = 0.3) +
  geom_smooth(se = FALSE) +
  labs(
    title = "Comparing sentence-level sentiment in two documents",
    x = "Sentence ID",
    y = "sentimentr value",
    colour = "Title"
  )
```

Questions to ask:

- Do the two texts show different sentiment patterns?
- Where are the peaks and dips?
- Which passages would you return to for close reading?

## Takeaway

Sentiment analysis can show broad affective patterns, but it is limited by:

- lexicon choice
- historical language
- genre and narrative voice
- negation
- irony
- the unit of analysis

Treat sentiment scores as prompts for interpretation, not final conclusions.
