---
title: "06. Bridge Exercise: From Thematic Neighbourhoods to Sentiment"
subtitle: "Text Analysis with R"
format:
  html:
    toc: true
    toc-depth: 2
    number-sections: true
execute:
  echo: true
  warning: false
  message: false
---

## Aim

This optional exercise connects two methods from the course:

1. use word vectors to build a thematic representation
2. find the words from that representation back in the corpus
3. analyse the sentiment of the sentences where those words appear

The goal is not to prove that a theme has one sentiment. The goal is to move from model output back to textual evidence.

```{r}
# List the packages used in this bridge exercise.
required_packages <- c("tidyverse", "text2vec", "tidytext")

# Check which packages are missing from this R installation.
missing_packages <- required_packages[!vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)]

# Install only packages that are missing.
if (length(missing_packages) > 0) {
  install.packages(missing_packages)
}

# Load packages for data manipulation, tokenisation, word vectors, and plotting.
library(tidyverse)
library(text2vec)
library(tidytext)
```

## Load the Prepared Data

We use three prepared objects:

- `corpus_docs`: full texts, used to train a small word-vector model
- `corpus_sentence`: sentence-level corpus, used to find sentences containing theme words
- `sentimentr_sent_values`: precomputed sentence-level sentiment scores

```{r}
# Load the document-level corpus used for word-vector training.
load("corpus_docs.RData")

# Load the sentence-level corpus used for retrieving examples.
load("corpus_sentence.RData")

# Load precomputed sentence sentiment values to avoid slow processing in class.
load("sentimentr_sent_values.RData")
```

```{r}
# Check the columns available in each object.
glimpse(corpus_docs)
glimpse(corpus_sentence)
glimpse(sentimentr_sent_values)
```

## Choose Your Theme

Fill in a small set of seed words for a theme you want to explore. Choose words that are likely to appear in the corpus.

```{r}
# Ready to fill: replace these seed words with a theme for your own research.
theme_seed_terms <- c("power", "king", "law", "rule")

# Ready to fill: give the theme a readable label for tables and plots.
theme_label <- "power and authority"

# Ready to fill: choose how many neighbouring words to keep.
n_neighbours <- 15
```

## Train a Small Word-Vector Model

This repeats the efficient workflow from the word-vector notebook. The sample size and model settings are modest so the exercise remains usable on Posit Cloud.

```{r}
# Choose how many documents to use for training.
# Reduce this number if Posit Cloud is slow.
n_documents <- 100

# Set a seed so the model is more reproducible.
set.seed(123)

# Prepare texts for modelling.
corpus_for_vectors <- corpus_docs %>%
  # Sort documents so the selected sample is stable.
  arrange(year, doc_id) %>%
  # Keep the first n_documents documents.
  slice_head(n = n_documents) %>%
  # Lowercase the text so differently capitalised words are treated alike.
  mutate(text = str_to_lower(text))
```

```{r}
# Split each document into word tokens.
tokens <- word_tokenizer(corpus_for_vectors$text)

# Create a text2vec iterator, which streams tokenised texts into later steps.
it <- itoken(
  tokens,
  ids = corpus_for_vectors$doc_id,
  progressbar = FALSE
)
```

```{r}
# Build a vocabulary and remove rare terms.
vocab <- create_vocabulary(it, stopwords = tidytext::stop_words$word) %>%
  # Keep terms that occur at least 20 times.
  prune_vocabulary(term_count_min = 20)

# Convert the vocabulary into a vectorizer for the matrix step.
vectorizer <- vocab_vectorizer(vocab)
```

```{r}
# Build a term co-occurrence matrix.
# The window size controls how far around each word the model looks for context.
tcm <- create_tcm(it, vectorizer, skip_grams_window = 5)
```

```{r}
# Train a small GloVe model.
glove <- GlobalVectors$new(rank = 50, x_max = 10)

# Fit the main word vectors.
main_vectors <- glove$fit_transform(tcm, n_iter = 8, convergence_tol = 0.01)

# Extract context vectors from the fitted model.
context_vectors <- glove$components

# Combine main and context vectors into the final word-vector matrix.
word_vectors <- main_vectors + t(context_vectors)
```

## Find a Thematic Neighbourhood

The helper function below finds words whose vectors are close to your seed terms.

```{r}
# Define a helper function for finding nearest words to one or more seed terms.
nearest_words <- function(target_terms, word_vectors, n = 20) {
  # Keep only seed terms that are present in the model vocabulary.
  available_terms <- intersect(target_terms, rownames(word_vectors))

  # Stop with a clear message if none of the seed terms are available.
  if (length(available_terms) == 0) {
    stop("None of the target terms are in the vocabulary. Try more common words.")
  }

  # Average the vectors for all available seed terms.
  target_vector <- colMeans(word_vectors[available_terms, , drop = FALSE])

  # Calculate cosine similarity between the theme vector and every word vector.
  similarities <- sim2(
    x = word_vectors,
    y = matrix(target_vector, nrow = 1),
    method = "cosine",
    norm = "l2"
  )[, 1]

  # Return the nearest words as a tidy table.
  tibble(
    word = names(similarities),
    similarity = as.numeric(similarities)
  ) %>%
    # Remove the original seed terms from the output.
    filter(!word %in% available_terms) %>%
    # Put the closest words first.
    arrange(desc(similarity)) %>%
    # Keep only the requested number of words.
    slice_head(n = n)
}
```

```{r}
# Check which seed terms are available in the model.
tibble(term = theme_seed_terms) %>%
  mutate(in_model = term %in% rownames(word_vectors))
```

```{r}
# Find words close to the seed terms.
theme_neighbours <- nearest_words(
  target_terms = theme_seed_terms,
  word_vectors = word_vectors,
  n = n_neighbours
)

# Inspect the thematic neighbourhood.
theme_neighbours
```

```{r}
# Plot the nearest words for the chosen theme.
theme_neighbours %>%
  mutate(word = fct_reorder(word, similarity)) %>%
  ggplot(aes(x = similarity, y = word)) +
  geom_col(fill = "steelblue") +
  labs(
    title = paste("Words near", theme_label),
    x = "Cosine similarity",
    y = NULL
  ) +
  theme_minimal()
```

## Find the Theme Back in the Corpus

Now we turn the model output into a search vocabulary. We combine the original seed terms with the nearest words and use them to retrieve sentences.

```{r}
# Combine the original seed words and the nearest words.
terms_to_trace <- c(theme_seed_terms, theme_neighbours$word) %>%
  # Keep each term only once.
  unique()

# Inspect the terms that will be searched in the corpus.
terms_to_trace
```

Some vector neighbours may be too general for the research question. This is not a problem to hide: it is part of interpretation. We can remove words that would retrieve too many irrelevant sentences.

```{r}
# Ready to fill: remove words that are too broad for your theme.
terms_to_remove <- c("lived", "true", "meet")

# Keep only the terms we want to trace back into the corpus.
terms_to_trace <- setdiff(terms_to_trace, terms_to_remove)

# Inspect the refined search vocabulary.
terms_to_trace
```

```{r}
# Build a regular expression that searches for any selected theme term.
# The word boundaries help avoid partial matches such as "law" inside "lawyer".
theme_pattern <- paste0("\\b(", str_c(terms_to_trace, collapse = "|"), ")\\b")
```

```{r}
# Retrieve sentences that contain at least one theme term.
theme_sentences <- corpus_sentence %>%
  # Create a lowercase copy of each sentence for matching.
  mutate(sentence_lower = str_to_lower(sentence)) %>%
  # Keep sentences where the theme pattern appears.
  filter(str_detect(sentence_lower, theme_pattern)) %>%
  # Extract the matched theme words from each sentence.
  mutate(matched_terms = str_extract_all(sentence_lower, theme_pattern)) %>%
  # Count how many theme terms appear in the sentence.
  mutate(n_matched_terms = lengths(matched_terms)) %>%
  # Turn the matched terms into a readable comma-separated string.
  mutate(matched_terms = map_chr(matched_terms, ~ str_c(unique(.x), collapse = ", "))) %>%
  # Remove the helper lowercase column.
  select(-sentence_lower)

# Inspect the first retrieved sentences.
theme_sentences %>%
  select(doc_id, title, author, year, sentence_id, matched_terms, sentence) %>%
  head(10)
```

```{r}
# Count how often each traced term appears in the retrieved sentences.
theme_term_counts <- theme_sentences %>%
  # Split the comma-separated matched terms back into one term per row.
  separate_rows(matched_terms, sep = ", ") %>%
  # Count term occurrences.
  count(matched_terms, sort = TRUE)

# Inspect the most common matched terms.
theme_term_counts
```

## Add Sentence Sentiment

We now join the retrieved sentences to the precomputed sentiment scores. This is faster than recalculating sentiment during the exercise.

```{r}
# Collapse the sentiment table to one row per document sentence.
# This avoids duplicated retrieved sentences if sentimentr produced more than one row.
sentence_sentiment_values <- sentimentr_sent_values %>%
  group_by(doc_id, sentence_id) %>%
  summarise(
    sentiment_value = mean(sentiment_value, na.rm = TRUE),
    .groups = "drop"
  )
```

```{r}
# Join the retrieved sentences to the precomputed sentiment values.
theme_sentence_sentiment <- theme_sentences %>%
  left_join(
    sentence_sentiment_values,
    by = c("doc_id", "sentence_id")
  )

# Inspect the retrieved sentences with sentiment values.
theme_sentence_sentiment %>%
  select(doc_id, title, author, year, sentence_id, matched_terms, sentiment_value, sentence) %>%
  head(10)
```

```{r}
# Summarise the sentiment of theme sentences by document.
theme_sentiment_by_document <- theme_sentence_sentiment %>%
  group_by(doc_id, title, author, year) %>%
  summarise(
    n_sentences = n(),
    average_sentiment = mean(sentiment_value, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(desc(n_sentences))

# Inspect the document-level summary.
theme_sentiment_by_document
```

```{r}
# Plot average sentiment for documents with enough retrieved sentences.
theme_sentiment_by_document %>%
  # Keep documents where the theme appears in at least five sentences.
  filter(n_sentences >= 5) %>%
  # Keep the 20 documents with most retrieved sentences.
  slice_max(n_sentences, n = 20) %>%
  # Order titles by average sentiment for the plot.
  mutate(title = fct_reorder(title, average_sentiment)) %>%
  ggplot(aes(x = average_sentiment, y = title, size = n_sentences)) +
  geom_point(alpha = 0.7) +
  labs(
    title = paste("Sentiment of sentences about", theme_label),
    x = "Average sentence sentiment",
    y = NULL,
    size = "Sentences"
  ) +
  theme_minimal()
```

## Return to Sentences

The plot is only a guide. The next step is to inspect actual sentences, especially the strongest positive and negative examples.

```{r}
# Ready to fill: choose how many example sentences to inspect.
n_examples <- 8

# Inspect the most negative retrieved sentences.
theme_sentence_sentiment %>%
  arrange(sentiment_value) %>%
  select(title, author, year, matched_terms, sentiment_value, sentence) %>%
  slice_head(n = n_examples)
```

```{r}
# Inspect the most positive retrieved sentences.
theme_sentence_sentiment %>%
  arrange(desc(sentiment_value)) %>%
  select(title, author, year, matched_terms, sentiment_value, sentence) %>%
  slice_head(n = n_examples)
```

## Guided Exercises

### Exercise 1: Change the Theme

```{r}
# Ready to fill: choose a new theme and rerun the notebook from "Choose Your Theme".
theme_seed_terms <- c("city", "street", "house", "room")
theme_label <- "space and place"
```

Questions to ask:

- Which words does the model add to your initial theme?
- Are those additions useful, surprising, or misleading?
- Which words would you remove before searching the corpus?

### Exercise 2: Compare Two Theme Vocabularies

```{r}
# Ready to fill: choose two themes to compare.
theme_a_terms <- c("city", "street", "house")
theme_b_terms <- c("country", "field", "garden")

# Find neighbours for both themes.
theme_comparison <- bind_rows(
  nearest_words(theme_a_terms, word_vectors, n = 10) %>%
    mutate(theme = "theme A"),
  nearest_words(theme_b_terms, word_vectors, n = 10) %>%
    mutate(theme = "theme B")
)

# Inspect both neighbourhoods.
theme_comparison
```

Questions to ask:

- Do the two themes retrieve different kinds of words?
- Would both themes be suitable for sentence-level sentiment analysis?
- What would count as meaningful evidence for your research question?

### Exercise 3: Interpret the Limits

```{r}
# Count how many retrieved sentences have neutral, negative, or positive sentiment.
theme_sentence_sentiment %>%
  mutate(
    sentiment_group = case_when(
      sentiment_value > 0 ~ "positive",
      sentiment_value < 0 ~ "negative",
      TRUE ~ "neutral"
    )
  ) %>%
  count(sentiment_group)
```

Questions to ask:

- Are there enough sentences to interpret the pattern?
- Are the theme words central to the sentence, or only incidental?
- Does the sentiment describe the theme, a character, an event, or the narrator's phrasing?

## Takeaway

This workflow joins three levels of analysis:

- word vectors suggest a thematic vocabulary
- corpus retrieval shows where that vocabulary appears
- sentence sentiment gives a broad affective pattern to investigate

The strongest interpretations come from moving between the table, the plot, and the sentences themselves.
